---
title: "Batch Effect Diagnostics"
format: html
---

In this section, we will use a sample data included in this package to provide a thorough introduction to the key functionalities of the ComBatFamQC package for batch effect diagnostics.

# Data
___

The sample dataset in the package is the longitudinal cortical thickness data from the
[Alzheimer’s Disease Neuroimaging Initiative](https://adni.loni.usc.edu/
) (ADNI) study (Jack et al. (2008)), which con-
tains cortical thickness measures of 62 regions from 663 unique participants with images
collected longitudinally between 2-6 visits. The processed data can be downloaded from [GitHub](
https://github.com/ntustison/CrossLong). 

The key variables of interest are presented below:

-   Batch variable: 
    -   <span style="color:blueviolet;">*manufac*</span>: MRI manufacturer
-   Covariates
    -   <span style="color:darkseagreen;">*timedays*</span>: time related variable
    -   <span style="color:darkseagreen;">*Age*</span>
    -   <span style="color:darkseagreen;">*SEX*</span> 
    -   <span style="color:darkseagreen;">*DIAGNOSIS*</span>: (AD, CN, LMCI)
    
    (We consider an interaction effect between *timedays* and *DIAGNOSIS*)
-   Features to harmonize
    -   62 <span style="color:coral;">cortical thickness</span> regions
-   Random Effect
    -   <span style="color:cornflowerblue;">*subid*</span> : Subject IDs.

# Interactive Batch Effect Diagnostics
___

Users can launch the Shiny app using the following code:

::: {.callout-note collapse="true"}
## R code
```{r}
features <- colnames(adni)[c(43:104)]
covariates <- c("timedays", "AGE", "SEX", "DIAGNOSIS")
interaction <- c("timedays,DIAGNOSIS")
batch <- "manufac"
result_lmer <- visual_prep(type = "lmer", features = features, batch = batch, covariates = covariates, interaction = interaction, smooth = NULL, random = "subid", df = adni)
comfam_shiny(result_lmer)
```
:::

Alternatively, users can utilize the command-line interface via **ComBatQC_CLI.R**. The essential steps have been streamlined, significantly simplifying the workflow and making it easier to get started. The following parameters must be assigned values:

-   `--diagnosis/-d`: TRUE (by default)
-   `--visualization/-v`: TRUE (by default)
-   `--features/-f`: position of features/rois data (column numbers)
-   `--covariates/-c`: column position of the covariates whose effects should be preserved
-   `--batch/-b`: column position of the batch variable
-   `--smooth/-s`:  column position of the smooth term (if a gam model is used)
-   `--random/-r`: column position of the random effect term (if a lmer model is used)

::: {.callout-tip collapse="true"}
## CLI code
```{bash, eval = FALSE}
Rscript path/to/combatQC_CLI.R path/to/unharmonized_data.csv -d TRUE -v TRUE --features 43-104 --covariates 9,11,13-14 -b 16 -i 9*14 -r 3 -m lmer 
```
:::

## Data Overview

::: {#fig-cern}

{{< video https://youtu.be/5OJU4oVZ6Ow >}}

ComBatFamQC: Data Overview

:::

## Summary

::: {#fig-cern}

{{< video https://youtu.be/cLI9WkCPLF8 >}}

ComBatFamQC: Summary

:::

## Residial Plot

::: {#fig-cern}

{{< video https://youtu.be/EME2-E3UEXI >}}

ComBatFamQC: Residual Plot

:::

## Diagnosis of Global Batch Effect

::: {#fig-cern}

{{< video https://youtu.be/-CDK0qx5_xI >}}

ComBatFamQC: Diagnosis of Global Batch Effect

:::

## Diagnosis of Individual Batch Effect

::: {#fig-cern}

{{< video https://youtu.be/UBTw3REUlSw >}}

ComBatFamQC: Diagnosis of Individual Batch Effect

:::

## Interactive Harmonization

::: {#fig-cern}

{{< video https://youtu.be/CyAxIY9Cagk >}}

ComBatFamQC: Interactive Harmonization

:::


# Export Batch Effect Diagnosis Result
___

Users can save the batch effect diagnosis results using the following code. There are two ways to export the results: 1) as a combined EXCEL file, or 2) as a Quarto report.

-   **Export to a combined EXCEL file**

::: {.callout-note collapse="true"}
## R code
```{r}
diag_save(path = "path/to/dir", result = result_lmer, use_quarto = FALSE)
```
:::

-   **Export to a Quarto Report**

::: {.callout-note collapse="true"}
## R code
```{r}
library(quarto)
diag_save(path = "path/to/dir", result = result_lmer, use_quarto = TRUE)
```
:::

Using the command-line interface requires users to set the following parameter:

-   `--visualization/-v`: FALSE
-   `--outdir`: Path to a directory for saving the batch effect diagnosis result
-   `--quarto/-q`: Whether to generate a Quarto report.

::: {.callout-tip collapse="true"}
## CLI code
```{bash, eval = FALSE}
Rscript path/to/combatQC_CLI.R path/to/unharmonized_data.csv -d TRUE -v FALSE --features 43-104 --covariates 9,11,13-14 -b 16 -i 9*14 -r 3 -m lmer --outdir /path/to/dir --quarto TRUE
```
:::